
---
title: "Small extreme bandits problem with known configuration"
output:
  pdf_document: default
  html_document:
    highlight: pygments
    theme: spacelab
---

```{r setup, echo =FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.pos = 'h', fig.align = 'center')
knitr::opts_chunk$set(fig.cap = "",  fig.path = "Plot")
library(coda)
library(evd)
library(coda)
```

* * *
### Problem setup

Settings 
- K = 2 arms (k= 1, 2)
- D = 3 sources (d = 1, 2, 3)
- arm k=1 connects to source d = 1, 2 ($a_{11} = 1, a_{12} = 1, a_{13} = 0$)
- arm k=2 connects to source d = 2, 3 ($a_{21} = 0, a_{22} = 1, a_{23} = 1$)
- shape parameter of the three arms are $\alpha_1 = 1, \alpha_2 = 5, \alpha_3 = 10$

```{r}
set.seed(1)
# parameter setup
K = 2
D = 3
A = matrix(c(1,1,0,0,1,1),nrow = K, ncol = D,byrow = TRUE)
alpha_truth = c(1,5,10)
```

Distribution of output from the three sources
```{r}
x = seq(0.1, 10, 0.1)
pdf = cbind(dfrechet(x, shape = alpha_truth[1]), 
      dfrechet(x, shape = alpha_truth[2]), dfrechet(x, shape = alpha_truth[3]))
matplot(x,pdf, type = 'l', ylab = 'pdf', xlab = 'value',
        col = c('black', 'green', 'red'), lty = 1:3, lwd = 1,
        main = 'output of different sources')
legend('right', legend = alpha_truth, col = c('black', 'green', 'red'), lty = 1:3, lwd = 1)
```

Calculate the likelihood from the history (not vectorized)
```{r}
arm_hist = c(1,2,1,2) # keep history of what arms were played 
Y_hist = c(3,4,5,6) # keep history of rewards

# evaluate likelihood at alpha_current
alpha_current = alpha_truth # current value of shape parameter in the chain
alpha_d = alpha_current
likelihood = 1
for (step in 1:length(arm_hist)) {
  It = arm_hist[step] # the arm played at that step
  yt = Y_hist[step] # the reward at that step
  a_Itd = A[It,] # the connection at that step
  polynomial_terms = sum(a_Itd*alpha_d*yt^(-alpha_d-1))
  exponential_terms = exp(-sum(a_Itd*yt^(-alpha_d)))
  likelihood = likelihood*polynomial_terms*exponential_terms
}
likelihood
```

Calculate the likelihood from the history (vectorized)
```{r}
arm_hist = c(1,2,1,2) # keep history of what arms were played
Y_hist = c(3,4,5,6) # keep history of rewards

# evaluate likelihood at alpha_current
alpha_current = alpha_truth # current value of shape parameter in the chain
It = arm_hist # a vector of size #times
yt = matrix(Y_hist, nrow = length(arm_hist), ncol = D, byrow = FALSE) # a matrix of size #times x #sources
a_Itd = A[It,] # a matrix of size #times x #sources
alpha_d = matrix(alpha_current, nrow = length(arm_hist), ncol = D, byrow = TRUE)
polynomial_terms = rowSums(a_Itd*alpha_d*yt^(-alpha_d-1)) # a vector of size #times
exponential_terms = exp(-rowSums(a_Itd*yt^(-alpha_d))) # a vector of size #times
likelihood = prod(polynomial_terms*exponential_terms)
likelihood
```

```{r}
compute_likelihood <- function(arm_hist, Y_hist, alpha, A){
  # Compute the likelihood of given alpha and structure A
  # arm_hist: a vector of length #times indicating which arms were played in history
  # Y_hist: a vector of length #times recording the rewards we received from playing arm_hist
  # alpha: a vector of length D #sources indicating the shape parameters of sources at which the likelihood will be evaluated
  # A: a matrix of size #arms x #sources indicating the current configuration of the problem
  # return: likelihood
  
  It = arm_hist # a vector of size #times
  yt = matrix(Y_hist, nrow = length(arm_hist), ncol = D, byrow = FALSE) # a matrix of size #times x #sources
  a_Itd = A[It,] # a matrix of size #times x #sources
  alpha_d = matrix(alpha, nrow = length(arm_hist), ncol = D, byrow = TRUE)
  polynomial_terms = rowSums(a_Itd*alpha_d*yt^(-alpha_d-1)) # a vector of size #times
  exponential_terms = exp(-rowSums(a_Itd*yt^(-alpha_d))) # a vector of size #times
  likelihood = prod(polynomial_terms*exponential_terms)
  return(likelihood)
}
```

```{r}
compute_loglikelihood <- function(arm_hist, Y_hist, alpha, A){
  # Compute the loglikelihood of given alpha and structure A
  # arm_hist: a vector of length #times indicating which arms were played in history
  # Y_hist: a vector of length #times recording the rewards we received from playing arm_hist
  # alpha: a vector of length D #sources indicating the shape parameters of sources at which the likelihood will be evaluated
  # A: a matrix of size #arms x #sources indicating the current configuration of the problem
  # return: likelihood
  
  It = arm_hist # a vector of size #times
  yt = matrix(Y_hist, nrow = length(arm_hist), ncol = D, byrow = FALSE) # a matrix of size #times x #sources
  a_Itd = A[It,] # a matrix of size #times x #sources
  alpha_d = matrix(alpha, nrow = length(arm_hist), ncol = D, byrow = TRUE)
  polynomial_terms = rowSums(a_Itd*alpha_d*yt^(-alpha_d-1)) # a vector of size #times
  exponential_terms = exp(-rowSums(a_Itd*yt^(-alpha_d))) # a vector of size #times
  loglikelihood = sum(log(polynomial_terms*exponential_terms))
  return(loglikelihood)
}
```

```{r}
compute_likelihood(arm_hist, Y_hist, alpha_current, A)
compute_loglikelihood(arm_hist, Y_hist, alpha_current, A)
```

compute acceptance ratio for metropolis-hasting algo

update alpha_current to alpha_new or not
The proposal distribution is gamma(alpha*tuning, tuning)
```{r}
# suppose we want to sample alpha of source 1
alpha_current = c(2,5,10)
alpha_new = c(1,5,10) # propose new value for source 1 shape parameter
d = 1 # index of the source that we are going to update the shape parameter
tuning = 0.5 # tuning parameter of the proposal distribution - control variance around the current iteration
```

```{r}
compute_acceptance_ratio <- function(arm_hist, Y_hist, alpha_current, alpha_new, d, A){
  # Compute the acceptance ratio of the proposed dth shape parameter in alpha_new with the current value in alpha_current under Metropolis-Hasting method
  # arm_hist: a vector of length #times indicating which arms were played in history
  # Y_hist: a vector of length #times recording the rewards we received from playing arm_hist
  # alpha_current: a vector of length D #sources indicating the current shape parameters of sources
  # alpha_new: a vector of length D #sources indicating the new shape parameters of sources
  # d: the index of the source that we are going to update
  # A: a matrix of size #arms x #sources indicating the current configuration of the problem
  # return: r: acceptance ratio for the newly proposed shape parameter of source #d
  
  a = 1 # prior parameters of alpha Gamma(a,b)
  b = 1 # prior parameters of alpha Gamma(a,b)
  
  # extract the shape parameter of the souce we want to update
  alpha_current_d = alpha_current[d]
  alpha_new_d = alpha_new[d]
  
  # compute the pdf of all parts in acceptance ratio formula for Metropolis-Hasting algo
  loglikelihood_alpha_current = compute_loglikelihood(arm_hist, Y_hist, alpha_current, A) #likelihood at alpha_current
  loglikelihood_alpha_new = compute_loglikelihood(arm_hist, Y_hist, alpha_new, A) #likelihood at alpha_new
  prior_alpha_current = (alpha_current_d^(a-1))*exp(-b*alpha_current_d) #prior at alpha_current
  prior_alpha_new = (alpha_new_d^(a-1))*exp(-b*alpha_new_d) #prior at alpha_new
  # g(alpha_current_d|alpha_new_d) Gamma(alpha_new_d*tuning,tuning)
  sample_alpha_current = dgamma(alpha_current_d, shape = alpha_new_d*tuning, rate = tuning)  
  # g(alpha_new_d|alpha_current_d) Gamma(alpha_current_d*tuning,tuning)
  sample_alpha_new = dgamma(alpha_new_d, shape = alpha_current_d*tuning, rate = tuning) 
  
  # compute acceptance ratio
  r = (exp(loglikelihood_alpha_new -loglikelihood_alpha_current)*prior_alpha_new*sample_alpha_current)/
  (prior_alpha_current*sample_alpha_new)

  return(r)
}
```

```{r}
compute_acceptance_ratio(arm_hist, Y_hist, alpha_current, alpha_new, d, A)
```

Try to infer alpha based on samples

Experiment 1: try arm 1 200 times and arm 2 200 times
```{r}
set.seed(1)
# play 1st arm 200 times and then 2nd arm 200 times
arm_hist = c(rep(1,200), rep(2,200))
Y_hist = rep(0, length(arm_hist))
# simulate data
for (step in 1:length(arm_hist)) {
  It = arm_hist[step] # arm that was played at that step
  active_source = (A[It,])*(1:D) # which sources are active
  Y_hist[step] = max(rfrechet(sum(A[It,]),shape = alpha_truth[active_source]))
}
```

```{r}
# start the chain with alpha = c(2,4,6)
alpha_current = c(2,4,6)
tuning = 5 # tuning parameter of the proposal distribution
burnin = 1000
size = 5000 # total number of posterior samples before thining
thining = 5
ALPHA = matrix(0, nrow = size, ncol = D) # matrix to save samples from posterior distribution
count = 0
accept = 0
for (t in 1:(burnin+size)) {
  for (d in 1:D) {
    alpha_current_d = alpha_current[d]
    # propose new values using proposal distribution Gamma(alpha_current_d*tuning,tuning)
    alpha_new_d = rgamma(1, shape = alpha_current_d*tuning, rate = tuning)
    alpha_new = alpha_current
    alpha_new[d] = alpha_new_d
    
    # compute acceptance ratio
    r = compute_acceptance_ratio(arm_hist, Y_hist, alpha_current, alpha_new, d, A)
    u = runif(1,0,1)
    count = count + 1
    if (u<r) {
      # accept proposed value
      alpha_current = alpha_new
      accept = accept + 1
    }
  }
  
  if (t>burnin) {
    # past burnin period already - save output
    ALPHA[t-burnin,] = alpha_current
  }
}
ALPHA = ALPHA[seq(1, size, thining),]
print(accept/count)
```

```{r}
ALPHA.mcmc <- mcmc(ALPHA, start = 1)
summary(ALPHA.mcmc)
plot(ALPHA.mcmc)
autocorr.plot(ALPHA.mcmc)
```

To do list
- Check current code for r (done)
- implement r as a function (done)
- try simple simulation plan (done)
- check tuning parameter for Metropolis (done)
- change likelihood to log likelihood (done)
- simulate (Thursday)
* * *



